A Wait-free Queue with Polylogarithmic Step Complexity

### 2022-12-29 podc_paper/algorithm.tex
working on code

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode}.
We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
We use \var{num\sub{deq}} similarly.
\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\Function{element}{FindResponse}{\var{int b, int i}}
    \State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
    \If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
        \State \linecomment{queue is empty when the dequeue is linearized}
        \State \Return \var{null} \label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State $\var{e} \assign\ \var{i} + \var{root.blocks[b-1].sum\sub{enq}} - 
	\var{root.blocks[b-1].size}$\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $\var{b}'\leq \var{b}$ with $\var{root.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $\var{i}' \assign\ \var{e} - \var{root.blocks}[\var{b}'-1].\var{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{root, $\var{b}'$, $\var{i}'$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

@@

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{25}

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
\While{\nf{start<end}}
\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
\If{\nf{blocks[mid].sum\sub{enq}<x}}
\State \nf{start \assign\ mid+1}
\Else
\State \nf{end \assign\ mid}
\EndIf
\EndWhile
\State\Return \nf{start}
\EndFunction{BinarySearch}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
\com\ the \nf{b}th block in the \nf{root}.}

\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
\State \tt{start \assign\ end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
\EndWhile \label{dsearchEnd}
\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DoublingSearch}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{94}

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
\State \tt{subblockIndex \assign left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
\Else
\State \tt{i \assign\ i-blocks[b].num\sub{enq-left}}
\State\tt{subblockIndex \assign\ right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
\EndIf
\EndFunction{GetEnqueue}

\end{algorithmic}
\end{algorithm}

>>>>

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
We use \var{num\sub{deq}} similarly.
\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

\Function{element}{FindResponse}{\var{int b, int i}}
    \State \linecomment{find response to \var{i}th dequeue in \var{root.blocks[b]}}
    \If{\var{root.blocks[b-1].\size + root.blocks[b].num\sub{enq}} $<$ \var{i}}\label{checkEmpty}
        \State \Return \var{null} \hfill \com\ queue is empty when dequeue occurs\label{returnNull}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State $\var{e} \assign\ \var{i} + \var{root.blocks[b-1].sum\sub{enq}} - 
	\var{root.blocks[b-1].size}$\label{computeE}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $\var{b}'\leq \var{b}$ with $\var{root.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{e}$
		\State \linecomment{find rank of enqueue within its block}
		\State $\var{i}' \assign\ \var{e} - \var{root.blocks}[\var{b}'-1].\var{sum\sub{enq}}$
        \State \Return \Call{GetEnqueue}{root, $\var{b}'$, $\var{i}'$}\label{findAnswer}
    \EndIf
\EndFunction{FindResponse}

@@

%\begin{algorithm}
%\caption{\tt{\sl{Node}}}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{25}
%
%\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
%\Statex \com\ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in\\
%\com\ \tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
%\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
%\While{\nf{start<end}}
%\State \tt{\sl{int}} \tt{mid \assign\ floor((start+end)/2)}
%\If{\nf{blocks[mid].sum\sub{enq}<x}}
%\State \nf{start \assign\ mid+1}
%\Else
%\State \nf{end \assign\ mid}
%\EndIf
%\EndWhile
%\State\Return \nf{start}
%\EndFunction{BinarySearch}
%
%\end{algorithmic}
%\end{algorithm}

%\begin{algorithm}
%\caption{\tt{\sl{Root}}}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{36}
%\Statex
%\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
%\Statex \com\ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})$ is $E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within \\
%\com\ the \nf{b}th block in the \nf{root}.}
%
%\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
%\State \tt{start \assign\ end-1} \label{dsearchStart}
%\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{e}}
%\State \tt{start \assign\ max(start-(end-start), 0)} \label{doubling}
%\EndWhile \label{dsearchEnd}
%\State \tt{b \assign\ root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
%\State \tt{i \assign\ e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
%\State\Return \tt{<b,i>}
%\EndFunction{DoublingSearch}
%\end{algorithmic}
%\end{algorithm}

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{101}

\Function{element}{GetEnqueue}{\var{Node v, int b, int i}} \Comment{returns argument of ith enqueue in $E(\var{v.blocks[b]})$}
    \State \linecomment{Precondition: $\var{i}\geq 1$ and $E(\var{v.blocks[b]})$ contains at least \var{i} enqueues}

    \If{\var{v} is a leaf node}
        \State\Return \var{blocks[b].element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v.left.blocks[v.blocks[b-1].\eleft].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v.right.blocks[v.blocks[b-1].\eright].sum\sub{enq}} \Comment{\#\ of enqueues in \var{v.blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \cmt{required enqueue is in \var{v.left}}
        	\State \linecomment{Find enqueue's block in \var{v.left} and its rank within block}
            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eleft}+1..\var{v.blocks[b].\eleft}] s.t. $\var{v.left.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{left}}$
            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.left.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{left}})$
            \State \Return\Call{GetEnqueue}{\var{v.left}, $\var{b}'$, $\var{i}'$} 
        \Else \Comment{required enqueue is in \var{v.right}}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        	\State \linecomment{Find enqueue's block in \var{v.right} and its rank within block}
            \State find min $\var{b}'$ in range [\var{v.blocks[b-1].\eright}+1..\var{v.blocks[b].\eright}] s.t. $\var{v.right.blocks}[\var{b}'].\var{sum\sub{enq}} \geq \var{i} + \var{prev\sub{right}}$
            \State $\var{i}'$ \assign\ $\var{i} - (\var{v.right.blocks}[\var{b}'-1].\var{sum\sub{enq}} - \var{prev\sub{right}})$
            \State \Return\Call{GetEnqueue}{\var{v.right}, $\var{b}'$, $\var{i}'$} 
        \EndIf
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued. \here{Would this routine be simpler to write if i parameter be the rank within whole node rather than within the block?}}
\end{figure}
